Measurement Bias from Address Aliasing

### 2014-05-27 paper.tex
Add related work reference
Remove comments
Big changes to the abstract.

% Motivation -- Problem statement -- Approach -- Results -- Conclusions
\begin{abstract} % Should be 150 -- 200 words max
Understanding program performance and obtaining accure measurements is important in performance analysis, but measurements can be \emph{biased} by external factors.
Predicting or accounting for measurment bias is challenging, as the interaction with low level hardware mechanisms is complex.  
In this paper we show how \emph{address aliasing} can be the underlying mechanism resulting in measurement bias on modern Intel microarchitectures.
Our approach is to use hardware performance counter to analyze performance data over controlled series of execution contexts.
We show how address aliasing can explain bias from two external factors: environment variable size, and characteristics of dynamically linked heap allocators.

The perfomance impact of unfavorable memory contexts can be significant, with as much as $2x$ speedup in one of our examples.
It also turns out the allocation pattern of common libraries tends to give \emph{worst case} alignment by default, by favoring page alignment large allocations.
While the aliasing mechanism itself is not new, we provide the connection to measurement bias and show how to use this knowledge not only to account for bias but also explicitly optimize for it.
Our results should provide motivation for more architecture specific optimization, for example in heap allocator heuristics.
\end{abstract}

@@

The rest of this paper is structured as follows: Our methodology and experimental setup is explained in the next section, then section \ref{sec:aliasing} introduces ``4K aliasing'', providing necessary background for interpreting the later results.
Section \ref{sec:environment} presents an analysis of how the size of environment variables causes bias.
In Section \ref{sec:heap} we go through a similar analysis for bias to heap address alignment.
Section \ref{sec:conclusions} includes a summary and conclusion.

@@

%\begin{figure}
  % \lstset{%
  %   basicstyle=\ttfamily\small\bfseries,
  %   frame=tb,
  %   xleftmargin=0.2\textwidth,
  %   xrightmargin=0.2\textwidth
  % }
  % [xleftmargin=0.2\textwidth,xrightmargin=0.2\textwidth]
  \begin{lstlisting}
    static int i, j, k;
    int main() {
        int g = 0, inc = 1;
        for (; g < 65536; g++) {
            i += inc;
            j += inc;
            k += inc; 
        }
        return 0;
    }
  \end{lstlisting}

  % Can have caption with figure environment
  %\caption{Micro-kernel succeptible to bias from variations to system environment variables.\label{lst:loopkernel}}
%\end{figure}

>>>>

% Motivation -- Problem statement -- Approach -- Results -- Conclusions
\begin{abstract} % Should be 150 -- 200 words max
%Understanding program performance and obtaining accure measurements is important in performance analysis, but measurements can be \emph{biased} by external factors.
%Predicting or accounting for measurment bias is challenging, as the interaction with low level hardware mechanisms is complex.  
%In this paper we show how \emph{address aliasing} can be the underlying mechanism resulting in measurement bias on modern Intel microarchitectures.
%Our approach is to use hardware performance counter to analyze performance data over controlled series of execution contexts.
%We show how address aliasing can explain bias from two external factors: environment variable size, and characteristics of dynamically linked heap allocators.

%The perfomance impact of unfavorable memory contexts can be significant, with as much as $2x$ speedup in one of our examples.
%It also turns out the allocation pattern of common libraries tends to give \emph{worst case} alignment by default, by favoring page alignment large allocations.
%While the aliasing mechanism itself is not new, we provide the connection to measurement bias and show how to use this knowledge not only to account for bias but also explicitly optimize for it.
%Our results should provide motivation for more architecture specific optimization, for example in heap allocator heuristics.
Understanding program performance and obtaining accurate measurements is important in performance analysis, but measurements can be \emph{biased} by external factors.
It has been shown that there can be significant performance differences based seemingly irrelevant factors like user name length.
In this paper we show an important underlying mechanism resulting in this type of measurement bias on modern Intel microarchitectures.

Our approach is to search for correlations between a large set of hardware performance counters and performance figures from multiple environment contexts.
We find that \emph{address aliasing} can explain bias from two external factors: environment variable size, and characteristics of dynamically linked heap allocators.

This clear result not only makes accounting and avoiding measurement bias much simpler, but enables explicit optimizations for it.
It also turns out the allocation pattern of common libraries tends to give \emph{worst case} alignment by default, by favoring page alignment large allocations.
The performance impact of unfavorable memory contexts can be significant, with as much as $2x$ speedup in one of our examples.
Our results should provide motivation for more architecture specific optimization, for example in heap allocator heuristics and compiler rules.

\end{abstract}

@@

The rest of this paper is structured as follows: Our methodology and experimental setup is explained in the next section, then section \ref{sec:aliasing} introduces ``4K aliasing'', providing necessary background for interpreting the later results.
Section \ref{sec:environment} presents an analysis of how the size of environment variables causes bias.
In Section \ref{sec:heap} we go through a similar analysis for bias to heap address alignment.
Related work is discussed in Section \ref{sec:related}.
Section \ref{sec:conclusions} includes a summary and conclusion.

@@

\begin{lstlisting}
    static int i, j, k;
    int main() {
        int g = 0, inc = 1;
        for (; g < 65536; g++) {
            i += inc;
            j += inc;
            k += inc; 
        }
        return 0;
    }
\end{lstlisting}
